Linting Linearity in Core/System FC

### 2023-09-22 language-v4/proofs/optimizations/BetaReduction.tex
Write on more optimisations

@@

\begin{tabbing}
% Using preservation
    (1) $\G; \D \vdash (\lambda \xl.~e)~e' : \vp$\\
    (2) $(\lambda \xl.~e) e' \longrightarrow e[e'/x]$\\
    (3) $\G; \D \vdash e[e'/x] : \vp$ \` by type preservation theorem (1,2)\\

@@

@@

>>>>

$\beta$-reduction evaluates a $\lambda$-abstraction application by substituting
the $\lambda$-bound variable $x$ with the argument $e'$ in the body of the
$\lambda$-abstraction $e$.
%
We consider two definitions of $\beta$-reductions, one that substitutes all
occurrences of a variable by an expression, as in call-by-name, and other which
creates a lazy let binding to share the result of computing the argument
expression amongst uses of the variable, as in call-by-need.

The first kind of $\beta$-reduction on term abstractions can be seen to
preserve linearity by case analysis. When the function is linear, the binding
$x$ is used exactly once in the body of the lambda, thus can be substituted by
an expression typed with linear resources, since the expression is guaranteed
to be used exactly once in place of $x$. The proof is direct by type preservation.

@@

\begin{tabbing}
% Using preservation
    (1) $\G; \D \vdash (\lambda \x[\pi].~e)~e' : \vp$\\
    (2) $(\lambda \x[\pi].~e) e' \longrightarrow e[e'/x]$\\
    (3) $\G; \D \vdash e[e'/x] : \vp$ \` by type preservation theorem (1,2)\\

@@

\noindent We assume the $\beta$-reduction with sharing (i.e. the one that creates a let
binding) is only applicable when the $\lambda$-abstraction has an unrestricted
function type. Otherwise, the call-by-name $\beta$-reduction is always
favourable, as we know the resource to be used exactly once and hence sharing
would be counterproductive, and result in an unnecessary heap allocation.
%
Consequently, the argument to the function must be unrestricted (hence use no
linear resources) for the term to be well-typed, and so it vacuously follows
that linearity is preserved by this transformation.

@@

\noindent Finally, $\beta$-reduction on multiplicity abstractions is also type
preserving. The argument of the application is a multiplicity rather than an
expression, so no resources are needed to type it, and since the body of the
lambda must treat the multiplicity $p$ as though it were a linear, the body
uses the argument linearly regardless of the instantiation of $p$ at $\pi$.

### 2023-09-22 language-v4/proofs/optimizations/CaseOfKnownConstructor.tex
Write and typeset multiple optimisation proofs
Write on more optimisations

@@

\begin{proof}
    Trivial through the progress theorem.
\end{proof}

>>>>

Case-of-known constructor is a transformation that essentially evaluates a case
expression of a known constructor at compile time, by substituting in the
matching alternative the case binder by the scrutinee and pattern variables by
constructor arguments.
%
Intuitively, either the case binder is used exactly once to consume the
resources of the scrutinee, or the pattern components whose matching
constructor argument uses linear resources are used exactly once, so one of the
substitutions is a no-op.  If the pattern variables are substituted by the
matching scrutinee expressions, the expressions are still only used once, and
if the case binder is substituted by the scrutinee, it is still used exactly
once.
%
The proof follows trivially from the preservation theorem.

@@

\begin{proof}~
\begin{tabbing}
    (1) $\G; \D, \D' \vdash \ccase{K~\ov{e}}{\z[\D][\s]~\{..., K~\ov{x} \to e_i\}} : \vp$\\
    (2) $\ccase{K~\ov{e}}{\z[\D][\s]~\{..., K~\ov{x} \to e_i\}} \longrightarrow e_i\ov{[e/x]}[K~\ov{e}/z]$\\
    (3) $\G; \D,\D' \vdash e_i\ov{[e/x]}[K~\ov{e}/z] : \vp$\` by preservation theorem (1,2)\\
\end{tabbing}
\end{proof}

### 2023-09-22 language-v4/proofs/optimizations/EtaConvs.tex
Write on more optimisations

>>>>

The $\eta$-conversion transformations are $\eta$-expansion and
$\eta$-reduction. In both transformations, linearity is preserved since the
resources used to type the function $f$ do not change neither when the lambda
and its argument are removed, nor when we add a lambda and apply $f$ to the
bound argument.

### 2023-09-22 language-v4/proofs/optimizations/Inlining.tex
Mre typesetting proofs
Write on more optimisations

@@

\begin{tabbing}
    (1) $\G,\G',\D \vdash \llet{\xD = e}{e'} : \vp$\\
    (2) $\G,\D,\xD \vdash e' : \vp$\` by inv. on (let)\\
    (3) $\G',\D \vdash e : \s$\` by inv. on (let)\\
    (4) $\hasnolinearvars{\G'}$\\
    (5) $\G,\G',\D\vdash e'[e/x] : \vp$\` by $\D$-subst. lemma (2,3)\\
    (6) $\G,\G',\D,\xD \vdash e'[e/x] : \vp$ \` by $Weaken_\Delta$\\
    (7) $\G,\G',\G',\D \vdash \llet{\xD = e}{e'[e/x]} : \vp$\` by (let) (7,3,4)\\
    (8) $\G,\G',\D \vdash \llet{\xD = e}{e'[e/x]} : \vp$\` by (4)\\
\end{tabbing}
\end{proof}

>>>>

Inlining substitutes occurrences of a let-bound $\D$-variable $x$
with the expression $e$ it is bound to, which determines its usage environment
$\D$. Intuitively, in the let body $e'$, $x$ can occur once or not at all: if
$x$ occurs, then the linear resources $\D$ used indirectly through $x$ are used
via $e$ instead; if $x$ does not occur, then the resources $\D$ are already
used linearly in $e'$ and the substitution is a no-op.

@@

\begin{tabbing}
    (1) $\G;\D,\D' \vdash \llet{\xD = e}{e'} : \vp$\\
    (2) $\G,\D \vdash e : \s$\` by inv. on (let)\\
    (3) $\G,\xD; \D, \D' \vdash e' : \vp$\` by inv. on (let)\\
    (4) $\G;\D,\D' \vdash e'[e/x] : \vp$\` by $\D$-subst. lemma (2,3)\\
    (5) $\G,\xD; \D,\D' \vdash e'[e/x] : \vp$ \` by (admissible) $Weaken_\Delta$\\
    (6) $\G;\D,\D' \vdash \llet{\xD = e}{e'[e/x]} : \vp$\` by (let) (2,5)\\
\end{tabbing}
\end{proof}
